White Dwarf Mergers on Adaptive Meshes\\ III. Mergers

### 2016-07-22 papers/wdmerger_II/paper.tex
Updates to papers

Given a set of nuclides and the reaction links between them, we now consider
how a burning step is performed in our software. The goal is to integrate the
vector ${\bm{Y}} = (Y_1, Y_2, \ldots, Y_n, e, T)$, where $Y_{n} = X_{n} / A_{n}$
is the molar fraction of species $n$, with $X_n$ the mass fraction and $A_n$ the
mass number of that species, $e$ is the energy released during the burn, and
$T$ is the temperature. The equation describing its evolution is given by
\begin{equation}
  \frac{d\bm{Y}}{dt} = f(\mathbf{Y}),
\end{equation}
where the components of the right-hand-side for the species come from the particular
nuclear burning network we are using. The energy release $e$ of the zone's burn
will change when the nuclear abundances evolve, according to
\begin{equation}
  \frac{\partial e}{\partial t} = N_A \sum_{n} \frac{\partial Y_{n}}{\partial t} m_{n} c^2,
\end{equation}
where $c$ is the speed of light and $m_n$ is the mass of each nuclide.

In a hydrostatic burn, we keep $\rho$ and $T$ fixed throughout, and use
the energy released at the end to compute a final temperature that is
thermodynamically consistent with the new internal energy. By contrast,
in a self-heating burn, we allow the temperature to evolve in response
to the burning (see\footnote{In the cited paper, a term based on the
thermodynamic chemical potential was included; we now believe
that it is incorrect to include such a term in the burn, since it
automatically sums to zero analytically.} \citet{maestro3}):
\begin{equation}
  \frac{dT}{dt} = \frac{1}{c_V}\frac{\partial e}{\partial t}
\end{equation}
(Although $T$ evolves during the burn so that the integration is physically
accurate, as for the hydrostatic method we discard the final value
for $T$ at the end of the burn and recompute a temperature for the zone that is
consistent with its new internal energy.) Here $c_V$ is the specific heat at
constant volume, which is provided by the equation of state.  During this burn,
we can keep $c_V$ constant using its initial value, or at each step we
can choose to re-evaluate the equation of state using the latest value of $(\rho, T)$.
The latter is more expensive but also more accurate, and we use it in this paper.
In practice we find that the cost is negligible in comparison to the more expensive
parts of the calculation, and it can significantly speed up convergence near NSE.
A third option presented by \citet{raskin:2010} is a so-called ``hybrid'' mode.
In this mode, by default we do a hydrostatic burn. If that burn fails, or if the net
energy change is negative, we do the burn again in self-heating mode. A final option
is a burn that limits the changes due to a burn to avoid numerically unstable burning.
This mode is discussed in \autoref{sec:unstable_burning} and we will call it a
``suppressed'' burn for the remainder of this work. All four options
are implemented in our burner software. The simulations shown in this work all
use the self-heating mode unless otherwise specified.

In our \microphysics\ repository we provide several software options for
solving a set of coupled stiff ODEs. For this work we use an implementation of
the well known variable-order Richardson extrapolation method presented by Stoer and
Bulirsch \cite{stoer:1980}, that is similar to the integrator which ships with
the original versions of the networks mentioned above. Previous work using the \boxlib\
codes typically used the classic \vode\ integrator \cite{vode}. We do maintain a version
of the software that is compatible with our software interfaces in the \microphysics\
repository, but we have largely shifted to integrators which are written in modern
Fortran as a consequence of our efforts to run our codes on GPUs. The Stoer and Bulirsch
integrator we provide satisfies this criterion; we also provide a rewrite of the VODE
BDF algorithm that uses modern Fortran features. The Stoer and Bulirsch algorithm
relies on a uniform relative error tolerance for all of the ODEs in the system, which
we set at $10^{-6}$ for all simulations described here.

@@

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
In a given timestep advance $\Delta t$, we first evolve the reactions alone through
a time interval $\Delta t / 2$. Then, we evolve the hydrodynamics for $\Delta t$,
and we evolve the reactions again for a further $\Delta t / 2$. The principal
drawback of this approach is that the reactions and the hydrodynamics can become
decoupled from each other. A common solution to this problem presented in
the literature has been to limit the size of the timestep and thereby limit the
extent of this decoupling \citep{raskin:2010,hawley:2012}, which we adopt here 
and have implemented in \castro. Defining the nuclear energy injection timescale 
$\tau_e$, and the species evolution timescale $\tau_{X_k}$,
\begin{align}
  \tau_e &\equiv \frac{e}{|\dot{e}|} \\
  \tau_{X_k} &\equiv \frac{X_k}{|\dot{X_k}|},
\end{align}
where $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning, and $\dot{X_k}$ is an estimate of the time rate of change 
of the mass fraction of the species with index $k$, we define burning-limited 
timesteps $\Delta t_{be}$ and $\Delta t_{bX_k}$:
\begin{align}
  \Delta t_{be} &= f_{e}\, \tau_e \label{eq:timestep_e}\\
  \Delta t_{bX_k} &= f_{X}\, \tau_{X_k}. \label{eq:timestep_X}
\end{align}
Given an estimate for $\dot{e}$, the factor $f_{e}$ determines by what 
fraction we would like to allow the internal energy to change
in the current timestep, under the assumption that $\dot{e}$ does not change from
timestep to timestep. Similarly, given an estimate for $\dot{X_k}$, the factor $f_{X}$ 
determines the maximum change in the mass fraction of any species. By making 
$f_{e}$ and $f_{X}$ smaller, we can control the magnitude of the decoupling 
between the reactions and the hydro. We choose by default $f_{X} = f_{e} = 0.1$, 
a choice which is slightly smaller than others in the literature. (Note that previous 
collision papers we are aware of only limited based on changes in internal energy, 
not on cahnges in species.)  The sensitivity of results to this choice will be discussed in 
\autoref{sec:parameters:timestepping}. The factors $f_{e}$ and $f_{X}$ can be set at runtime in \castro.

@@

To understand the consequences of this choice, and more broadly to 
understand the limitations of Strang splitting, we consider the 
basic outline of a single-level advance in an advection-reaction system:
\begin{enumerate}
  \item Evaluate timestep $\Delta t$ for the current advance
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Advance the hydrodynamics by $\Delta t$
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Return to Step 1
\end{enumerate}
Now, consider that during a head-on collision, initial nuclear burning 
will occur at the contact point between the two stars. Because of 
the staggered updates from splitting, the evolution effectively progresses 
as a cycle between burning for $\Delta t$ and getting fresh material 
advected into the contact point by the hydro update for $\Delta t$. 
When the collision begins, $\Delta t$ is controlled by the hydrodynamic 
stability criterion, and may be large enough that it is possible for 
the burning advance in Step 4 to completely burn the freshly advected 
material all the way to NSE. Consequently the evolution is no longer 
a good approximation to smooth burning of the in-falling material but
rather separate discrete burning and hydro steps, and the nature of 
the burning evolution will be quite different. Furthermore, by the 
time we return to Step 1 and estimate the next timestep size, all 
of the burning rates will be small again, and the instantaneous 
timestep limiter of mode 1 may actually substantially overestimate 
the needed timestep. The other modes will see that the energy/species  
substantially changed over the last timestep, but will still 
overestimate the needed timestep because the burning was quiescent
for at least some portion of the last advance. In particular, our 
experience has shown that using only the energy change criterion
is particularly susceptible to this phenomenon; silicon-group 
material can build up without changing the internal energy by a 
large fraction, so the timestep limiter is never triggered, and 
then in a single step a substantial amount of iron-group elements 
can be be generated. This can have non-trivial effects on the 
total amount of iron-group elements generated over the course of
the simulation. We have found that the addition of the 
limiter based on changes in species functionally precludes this.

With the timestep limited the way we advocate in this paper, 
the timesteps are generally short enough so that the errors 
due to splitting are small. Other approaches to the coupling 
between reactions and hydrodynamics have been proposed in the 
broader literature, especially iterative methods such as 
deferred corrections that allow each of these operators to 
feel the lagged effects of the other operators. For example,
in the context of low Mach number flows, \cite{nonaka:2012} have
used the method of spectral deferred corrections \citep{SDC} to
couple their advection-diffusion-reaction equation set. In our
context this would involve using the advection as a source term
for the burning network, and symmetrically including a source
term to the hydrodynamics corresponding to the energy release
from the burning. We are presently investigating such a method,
and it may form the basis of further work on this subject.

@@

One other point worth noting for the coupling of the reactions
and the hydrodynamics relates to the dual-energy formalism used
by \castro\ and many other hydrodynamics codes \cite{bryan:1995}.
\castro\ evolves separately equations for the total energy $(\rho E)$ and
for the internal energy $(\rho e)$. The former is conservative while the latter
is not, so for accurate hydrodynamics we prefer to use the total
energy when possible and to use the evolved internal energy variable
only in situations where the kinetic energy dominates the contribution
to the total energy. Indeed, for the cosmological purpose for which
this was originally developed, \cite{ENZO} choose a value $\eta_2 = 0.1$
so that $e$ is only updated to be equal to $E - \mathbf{v}^2/2$ if
$e > \eta_2\, E$. In other cases the evolved value of the internal
energy variable is preserved. However, this choice does not work
for our application, because the ultimate cause of the nuclear
burning in a white dwarf collision is the rapid conversion of
kinetic energy from the white dwarf bulk motion to thermal energy
as the white dwarfs slam into each other. Keeping $\eta_2$ large
prevents this from happening, and consequently the temperature
will never reach values high enough to generate significant
amounts of nickel. For this paper we choose $\eta_2 = 10^{-4}$,
which is low enough for the correct conversion of kinetic energy
but not so low that we need to be concerned about roundoff issues
caused by subtracting kinetic from total energy. This is also
the value that was used by \cite{hawley:2012}.

@@

\citet{kushnir:2013} point out that an inappropriate timestep is 
not the only way for the numerical discretization to cause 
severe errors in the burning. Another failure mode is when
the energy injection timescale
$t_e$ is shorter than the sound-crossing time $t_s$ in a zone.
When the sound-crossing time is too long, energy is built up in
a zone faster than it can be advected away by pressure waves.
This is obviously a problem inherent to numerically discretized
systems as the underlying fluid equations are continuous.
This can lead to a numerically seeded detonation caused by the
temperature building up too quickly in the zone; the detonation
is spurious in this case and should be avoided if possible.
The goal is to ensure that the following condition holds:
\begin{equation}
  t_s \leq f_{s} t_e \label{eq:burning_limiter_2}
\end{equation}
The sound crossing time, $t_s$, is given by $\Delta x / c_s$, 
where $c_s$ is the sound speed and $\Delta x$ is the (minimum) 
zone width. The parameter $f_{s}$ then determines the minimum
ratio of the nuclear energy generation timescale to the 
sound-crossing time. \citet{kushnir:2013} choose $f_{s} = 0.1$ 
for their simulations, and we do too (this parameter can be set 
at runtime in \castro).

\citet{kushnir:2013} implemented this criterion by artificially 
limiting the magnitude of the energy release after a burn. We
too have developed an option for our burner to do this,
the ``suppressed'' burning mode. In a suppressed burn, we limit
the changes to the state so that \autoref{eq:burning_limiter_2}
is always satisfied. To achieve this we directly multiply the
right-hand-side vector in the integration by a constant factor $F$
for all variables, where $F$ is the multiplicative factor needed to
be applied to $\dot{e}$ such that the equality in \autoref{eq:burning_limiter_2}
holds. (If the inequality is already satisfied, then the integration
vector is not modified.) We fix $t_s$ to be the value of the sound
crossing time at the beginning of the burn (that is, we do not
update it as the sound speed changes) and we fix the energy $e$
that goes into the estimate for $t_e$ to be the value of the
internal energy of the zone at the beginning of the burn. If
instead one allowed $c_s$ and $e$ to evolve with the burn, one
would obtain a less conservative limiter in the case of explosive
burning, as $c_s$ and $e$ are both increasing in this case.
As the point of the limiter is to ensure that the changes to the
\textit{original} energy are small enough so that the following
hydrodynamics update can advect away newly generated energy
quickly enough to avoid a nuerically seeded detonation,
we desire the most conservative version of the limiter. We discuss
our results with the suppressed burning mode in \autoref{sec:parameters:burningmode}.
Note that we have found that with this option enabled, it typically takes
many more timesteps to complete a burn than in self-heating mode.

Regardless of whether the suppressed burning mode works for this
particular problem, it is not physical, so we include for
consideration a different approach. If we insist that we cannot
cannot directly control the energy injection timescale, we 
must find a way to alter the sound-crossing timescale. 
We can achieve this by adding levels of refinement in 
regions that do not satisfy \autoref{eq:burning_limiter_2},
which effectively lowers the $\Delta x$ and thus the
sound-crossing time. We keep tagging zones for refinement
based on this criterion until the criterion is satisfied
on the finest level. Since the concern is regions that 
may detonate, we also tag nearby zones in a buffer region
which do not themselves satisfy the criterion,
so that a detonation in a single timestep cannot 
escape into non-refined regions. The width of the buffer 
region should thus be at least as large as the number of 
timesteps before a regridding procedure is performed.
We choose a value of two for both the number of zones in the 
buffer region and the number of steps in between regrids,
for all simulations in this paper.

We agree with \citet{kushnir:2013} that solving this numerical
instability is crucial to avoiding unphysical detonations.
A simulation that does not solve this problem will not obtain
the correct amount and will not converge properly with resolution.
This may justify the addition of many AMR levels to the domain
if a correct evaluation of the burning phase is desired.
Even if one cannot afford the full resolution required by this
AMR criterion, and chooses to limit the number of levels to some
predetermined maximum based on a constraint of computing time,
the added resolution will still go to the most-needed places.
However, a significant drawback to this approach is that it
does not turn on until after an unstable burning region has
been generated, so this AMR criterion cannot help in the case
where a spurious detonation begins in a single timestep on the coarse
grid, though it will kick in immediately after that timestep
to resolve the regions where the detonation has occurred.
A possible remedy to be explored in future work is to add
refinement pre-emptively in regions where the ratio $t_s / t_e$
approaches $f_{s}$ but does not yet exceed it. Another
possibility for this case would be to track how many levels
would have been needed to prevent the numerical detonation,
based on the stability criterion, and reject the timestep and
retry it from the starting state using that many levels of refinement.
The ability to reject a timestep and retry it is functionality that
we have recently added to \castro, so we may try this in future work.

@@

We implement the white dwarf collision problem in \castro\ using our \wdmerger\
package. We load the grid in the way most previous papers have done so:
the white dwarf centers of mass are initially separated by a distance of four times
the (secondary) white dwarf radius. Note that the radius of the white dwarf, and
consequently the initial distance, depends on the equation of state used, and this
can vary somewhat depending on the relevant physics used, such as Coulomb corrections
in the Helmholtz equation of state. However, the results do not really depend on the
exact initial distance as long as there is enough time for the WDs to distort in
response to tidal forces as they approach. Their initial velocity is that of
two point-masses in free-fall towards each other coming in from infinity, such that
the contact point is at the origin, and they approach each other on the $x$-axis.
The composition of the WDs is as described in Paper I: low-mass WDs are helium-dominated,
intermediate-mass WDs are carbon and oxygen dominated, and high-mass WDs contain
oxygen, neon, and magnesium. By default the WDs approach each other head-on, with
their respective centers lying on the $x$-axis, but we have also added an option
to have the WDs approach each other at a non-zero impact parameter $b$, where the
offset is perpendicular to the $x$-axis in the $y$ direction. We measure
$b$ in units of the secondary WD's radius, so that the default head-on case
corresponds to $b = 0$ and the case where the two WDs just graze past each other
corresponds to $b = 1$ (in practice it happens slightly differently because
of the WDs responding to each other's gravitational fields).

@@

Paper I established our methodology for calculating the gravitational wave signature
of a white dwarf merger using the post-Newtonian framework of \citet{blanchet:1990}.
While a typical merger event has a characteristic profile including a chirp and ringdown,
white dwarf collisions are noticeably distinct in their gravitational wave signatures, and
hence may be interesting targets for observation for future low frequency gravitational
wave detectors \citep{eLISA}. The strain increases monotonically for a collision until impact,
instead of oscillating back and forth at a characteristic frequency as for mergers.
The main caveat for these events compared to events
such as neutron star mergers is that due to the event being inherently less energetic,
the typical distance scale on which we could observe a white dwarf merger or collision
is the size of our own galaxy, with extragalactic events being unlikely to be
detected, at least with the first generation of space-based gravitational wave
detectors.

@@

>>>>

Given a set of nuclides and the reaction links between them, we now consider
how a burning step is performed in our software. The goal is to integrate the
vector ${\bm{Y}} = (Y_1, Y_2, \ldots, Y_n, e, T)$, where $Y_{n} = X_{n} / A_{n}$
is the molar fraction of species $n$, with $X_n$ the mass fraction and $A_n$ the
mass number of that species, $e$ is the energy released during the burn, and
$T$ is the temperature. The equation describing its evolution is given by
\begin{equation}
  \frac{d\bm{Y}}{dt} = f(\mathbf{Y}),
\end{equation}
where the components of the right-hand-side for the species come from the particular
nuclear burning network we are using. The energy $e$ of the zone
will change when the nuclear abundances evolve, according to
\begin{equation}
  \frac{\partial e}{\partial t} = N_A \sum_{n} \frac{\partial Y_{n}}{\partial t} m_{n} c^2,
\end{equation}
where $c$ is the speed of light and $m_n$ is the mass of each nuclide.

We define several burning modes that determine how $T$ and $e$ are evolved
during a nuclear burn. In a hydrostatic burn, which we call burning mode 0,
we keep $\rho$ and $T$ fixed throughout, and use 
the energy released at the end to compute a final temperature that is
thermodynamically consistent with the new internal energy. By contrast,
in a self-heating burn (mode 1), we allow the temperature to evolve in response
to the burning (see\footnote{In the cited paper, a term based on the
thermodynamic chemical potential was included; we now believe
that it is incorrect to include such a term in the burn, since it
automatically sums to zero analytically.} \citet{maestro3}):
\begin{equation}
  \frac{dT}{dt} = \frac{1}{c_V}\frac{\partial e}{\partial t}
\end{equation}
(Although $T$ evolves during the burn so that the integration is physically
accurate, as in the hydrostatic method we discard the final value
for $T$ at the end of the burn and recompute a temperature for the zone that is
consistent with its new internal energy.) Here $c_V$ is the specific heat at
constant volume, which is provided by the equation of state.  During this burn,
we can keep $c_V$ constant using its initial value, or at each step we
can choose to re-evaluate the equation of state using the latest value of $(\rho, T)$.
The latter is more expensive but also more accurate, and we use it in this paper.
In practice we find that the cost is small in comparison to the more expensive
parts of the calculation, and it can significantly speed up convergence near NSE.
A third option (mode 2) presented by \citet{raskin:2010} is a so-called ``hybrid'' mode.
In this mode, by default we do a hydrostatic burn. If that burn fails, or if the net
energy change is negative, we do the burn again in self-heating mode. A final option (mode 3)
is a burn that limits the changes due to a burn to avoid numerically unstable burning.
This mode is discussed in \autoref{sec:unstable_burning} and we will call it a
``suppressed'' burn for the remainder of this work. All four options
are implemented in our burner software. The simulations shown in this work all
use the self-heating mode unless otherwise specified.

In our \microphysics\ repository we provide several software options for
solving a set of coupled stiff ODEs. For this work we use an implementation of
the well known variable-order Richardson extrapolation method presented by \citet{stoer:1980},
that is similar to the integrator which ships with
the original versions of the networks mentioned above. Previous work using the \boxlib\
codes typically used the classic \vode\ integrator \citep{vode}. We do maintain a version
of the software that is compatible with our software interfaces in the \microphysics\
repository, but we have largely shifted to integrators which are written in modern
Fortran as a consequence of our efforts to run our codes on GPUs. The Stoer and Bulirsch
integrator we provide satisfies this criterion; we also provide a rewrite of the VODE
BDF algorithm that uses modern Fortran features. The Stoer and Bulirsch algorithm
relies on a uniform relative error tolerance for all of the ODEs in the system, which
we set at $10^{-6}$ for all simulations described here.

@@

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
In a given timestep advance $\Delta t$, we first evolve the reactions alone through
a time interval $\Delta t / 2$. Then, we evolve the hydrodynamics for $\Delta t$,
and we evolve the reactions again for a further $\Delta t / 2$. The principal
drawback of this approach is that the reactions and the hydrodynamics can become
decoupled from each other. A common solution to this problem presented in
the literature has been to limit the size of the timestep and thereby limit the
extent of this decoupling \citep{raskin:2010,hawley:2012}, which we adopt here 
and have implemented in \castro. Defining the nuclear energy injection timescale 
$\tau_e$, and the species evolution timescale $\tau_{X_k}$,
\begin{align}
  \tau_e &\equiv \frac{e}{|\dot{e}|} \\
  \tau_{X_k} &\equiv \frac{X_k}{|\dot{X_k}|},
\end{align}
where $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning, and $\dot{X_k}$ is an estimate of the time rate of change 
of the mass fraction of the species with index $k$, we define burning-limited 
timesteps $\Delta t_{be}$ and $\Delta t_{bX_k}$:
\begin{align}
  \Delta t_{be} &= f_{e}\, \tau_e \label{eq:timestep_e}\\
  \Delta t_{bX_k} &= f_{X}\, \tau_{X_k}. \label{eq:timestep_X}
\end{align}
Given an estimate for $\dot{e}$, the factor $f_{e}$ determines by what 
fraction we would like to allow the internal energy to change
in the current timestep, under the assumption that $\dot{e}$ does not change from
timestep to timestep. Similarly, given an estimate for $\dot{X_k}$, the factor $f_{X}$ 
determines the maximum change in the mass fraction of any species. By making 
$f_{e}$ and $f_{X}$ smaller, we can control the magnitude of the decoupling 
between the reactions and the hydro. A typical choice for $f_e$ parameter in the
literature is in the range of 0.2 or 0.3, while to our knowledge a limiter based on
$f_X$ has not been used by others performing these types of calculations. The sensitivity
of results to the value of these timestep limiters will be discussed in 
\autoref{sec:parameters:timestepping}. The factors $f_{e}$ and $f_{X}$ can be set at runtime in \castro.

@@

To understand the consequences of this choice, and more broadly to 
understand the limitations of Strang splitting, we consider the 
basic outline of a single-level advance in an advection-reaction system:
\begin{enumerate}
  \item Evaluate timestep $\Delta t$ for the current advance
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Advance the hydrodynamics by $\Delta t$
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Return to Step 1
\end{enumerate}
Now, consider that during a head-on collision, initial nuclear burning 
will occur at the contact point between the two stars. Because of 
the staggered updates from splitting, the evolution effectively progresses 
as a cycle between burning for $\Delta t$ and getting fresh material 
advected into the contact point by the hydro update for $\Delta t$. 
When the collision begins, $\Delta t$ is controlled by the hydrodynamic 
stability criterion, and may be large enough that it is possible for 
the burning advance in Step 4 to completely burn the freshly advected 
material all the way to NSE. Consequently the evolution is no longer 
a good approximation to smooth burning of the in-falling material but
rather separate discrete burning and hydro steps, and the nature of 
the burning evolution will be quite different. Furthermore, by the 
time we return to Step 1 and estimate the next timestep size, all 
of the burning rates will be small again, and the instantaneous 
timestep limiter of mode 1 may actually substantially overestimate 
the needed timestep. The other modes will see that the energy/species  
substantially changed over the last timestep, but will still
overestimate the needed timestep because the burning was quiescent
for at least some portion of the last advance. Our experience has
shown that none of these methods is flawless, and that limiting
based on only changes in internal energy is particularly susceptible
to this staggered burning phenomenon; silicon-group material can
build up without changing the internal energy by a  large fraction,
so the timestep limiter is never triggered, and then in a single step
a substantial amount of iron-group elements can be be generated,
perhaps forming a detonation. This can have non-trivial effects
on the total amount of iron-group elements generated over the course of
the simulation. We have found that the addition of the 
limiter based on changes in species functionally precludes this, and
so the two methods can complement each other.

With the timestep limited the way we advocate in this paper, 
the timesteps are generally short enough so that the errors 
due to splitting are small. Other approaches to the coupling 
between reactions and hydrodynamics have been proposed in the 
broader literature, especially iterative methods such as 
deferred corrections that allow each of these operators to 
feel the lagged effects of the other operators. For example,
in the context of low Mach number flows, \cite{nonaka:2012} have
used the method of spectral deferred corrections \citep{SDC} to
couple their advection-diffusion-reaction equation set. In our
context this would involve treating the full evolution equation
for each of the state variables as an ODE with a directly coupled
burning source term integrated at high order; the advective
flux is evaluated at the standard second order accuracy
and is included as an ODE source term.
We are presently investigating such a method,
and it may form the basis of further work on this subject.

@@

One other point worth noting for the coupling of the reactions
and the hydrodynamics relates to the dual-energy formalism used
by \castro\ and many other hydrodynamics codes \cite{bryan:1995}.
\castro\ evolves separately equations for the total energy $(\rho E)$ and
for the internal energy $(\rho e)$. The former is conservative while the latter
is not, so for accurate hydrodynamics we prefer to use the total
energy when possible and to use the evolved internal energy variable
only in situations where the kinetic energy dominates the contribution
to the total energy. Indeed, for the cosmological purpose for which
this was originally developed, \cite{ENZO} choose a value $\eta_2 = 0.1$
so that $e$ is only updated to be equal to $E - \mathbf{v}^2/2$ if
$e > \eta_2\, E$. In other cases the evolved value of the internal
energy variable is preserved. However, this choice is somewhat unsafe
for our application, because the ultimate cause of the nuclear
burning in a white dwarf collision is the rapid conversion of
kinetic energy from the white dwarf bulk motion to thermal energy
as the white dwarfs slam into each other. Keeping $\eta_2$ large
prevents this from happening, and consequently the temperature
will never reach values high enough to generate significant
amounts of nickel. For this paper we choose $\eta_2 = 10^{-4}$,
which is low enough for the correct conversion of kinetic energy
but not so low that we need to be concerned about roundoff issues
caused by subtracting kinetic from total energy. This is also
the value that was used by \cite{hawley:2012}.

@@

\citet{kushnir:2013} point out that an inappropriate timestep is 
not the only way for the numerical discretization to cause 
severe errors in the burning. Another failure mode is when
the energy injection timescale
$\tau_e$ is shorter than the sound-crossing time $\tau_s$ in a zone.
When the sound-crossing time is too long, energy is built up in
a zone faster than it can be advected away by pressure waves.
This is obviously a problem inherent to numerically discretized
systems as the underlying fluid equations are continuous.
This can lead to a numerically seeded detonation caused by the
temperature building up too quickly in the zone; the detonation
is spurious in this case and should be avoided if possible.
The goal is to ensure that the following condition holds:
\begin{equation}
  \tau_s \leq f_{s}\, \tau_e \label{eq:burning_limiter_2}
\end{equation}
The sound crossing time, $\tau_s$, is given by $\Delta x / c_s$, 
where $c_s$ is the sound speed and $\Delta x$ is the (minimum) 
zone width. The parameter $f_{s}$ then determines the minimum
ratio of the nuclear energy generation timescale to the 
sound-crossing time. \citet{kushnir:2013} choose $f_{s} = 0.1$ 
for their simulations, and we do too (this parameter can be set 
at runtime in \castro).

\citet{kushnir:2013} implemented this criterion by artificially 
limiting the magnitude of the energy release after a burn. We
too have developed an option for our burner to do this,
the ``suppressed'' burning mode. In a suppressed burn, we limit
the changes to the state so that \autoref{eq:burning_limiter_2}
is always satisfied. To achieve this we directly multiply the
right-hand-side vector in the integration by a constant factor $F$
for all variables, where $F$ is the multiplicative factor needed to
be applied to $\dot{e}$ such that the equality in \autoref{eq:burning_limiter_2}
holds. (If the inequality is already satisfied, then the integration
vector is not modified.) We fix $\tau_s$ to be the value of the sound
crossing time at the beginning of the burn (that is, we do not
update it as the sound speed changes) and we fix the energy $e$
that goes into the estimate for $\tau_e$ to be the value of the
internal energy of the zone at the beginning of the burn. If
instead one allowed $c_s$ and $e$ to evolve with the burn, one
would obtain a less conservative limiter in the case of explosive
burning, as $c_s$ and $e$ are both increasing in this case.
As the point of the limiter is to ensure that the changes to the
\textit{original} energy are small enough so that the following
hydrodynamics update can advect away newly generated energy
quickly enough to avoid a numerically seeded detonation,
we desire the most conservative version of the limiter. We discuss
our results with the suppressed burning mode in \autoref{sec:parameters:burningmode}.
Note that we have found that with this option enabled, it typically takes
many more timesteps to complete a burn than in self-heating mode.

Regardless of whether the suppressed burning mode works for this
particular problem, it is not physical, so we include for
consideration a different approach. If we insist that we
cannot directly control the energy injection timescale, we 
must find a way to alter the sound-crossing timescale. 
We can achieve this by adding levels of refinement in 
regions that do not satisfy \autoref{eq:burning_limiter_2},
which effectively lowers $\Delta x$ and thus the
sound-crossing time. We keep tagging zones for refinement
based on this criterion until the criterion is satisfied
on the finest level. Since the concern is regions that 
may detonate, we also tag nearby zones in a buffer region
which do not themselves satisfy the criterion,
so that a detonation in a single timestep cannot 
escape into non-refined regions. The width of the buffer 
region should thus be at least as large as the number of 
timesteps before a regridding procedure is performed.
We choose a value of two for both the number of zones in the 
buffer region and the number of steps in between regrids,
for all simulations in this paper.

We agree with \citet{kushnir:2013} that solving this numerical
instability is crucial to avoiding unphysical detonations.
A simulation that does not solve this problem will not obtain
the correct amount and will not converge properly with resolution.
This may justify the addition of many AMR levels to the domain
if a correct evaluation of the burning phase is desired.
Even if one cannot afford the full resolution required by this
AMR criterion, and chooses to limit the number of levels to some
predetermined maximum based on a constraint of computing time,
the added resolution will still go to the most-needed places.
However, a significant drawback to this approach is that it
does not turn on until after an unstable burning region has
been generated, so this AMR criterion cannot help in the case
where a spurious detonation begins in a single timestep on the coarse
grid, though it will kick in immediately after that timestep
to resolve the regions where the detonation has occurred.
A possible remedy to be explored in future work is to add
refinement pre-emptively in regions where the ratio $\tau_s / \tau_e$
approaches $f_{s}$ but does not yet exceed it. Another
possibility for this case would be to track how many levels
would have been needed to prevent the numerical detonation,
based on the stability criterion, and reject the timestep and
retry it from the starting state using that many levels of refinement.
As mentioned in \autoref{sec:burnerhydrocoupling}, the ability to reject
a timestep and retry it is functionality that
we have recently added to \castro, so we may try this in future work.

@@

We implement the white dwarf collision problem in \castro\ using our \wdmerger\
package. We load the grid in the way most previous papers have done so:
the white dwarf centers of mass are initially separated by a distance of four times
the (secondary) white dwarf radius. Note that the radius of the white dwarf, and
consequently the initial distance, depends on the equation of state used, and this
can vary somewhat depending on the relevant physics used, such as Coulomb corrections
in the Helmholtz equation of state. However, the results do not really depend on the
exact initial distance as long as there is enough time for the WDs to distort in
response to tidal forces as they approach. Their initial velocity is that of
two point-masses in free-fall towards each other coming in from infinity, such that
the contact point is at the origin, and they approach each other along the $x$-axis.
The composition of the WDs is as described in Paper I: low-mass WDs are helium-dominated,
intermediate-mass WDs are carbon and oxygen dominated, and high-mass WDs contain
oxygen, neon, and magnesium. By default the WDs approach each other head-on, with
their respective centers lying on the $x$-axis, but we have also added an option
to have the WDs approach each other at a non-zero impact parameter $b$, where the
offset is perpendicular to the $x$-axis in the $y$ direction. We measure
$b$ in units of the secondary WD's radius, so that the default head-on case
corresponds to $b = 0$ and the case where the two WDs just graze past each other
corresponds to $b = 1$ (in practice it happens slightly differently because
of the WDs responding to each other's gravitational fields).

@@

Paper I established our methodology for calculating the gravitational wave signature
of a white dwarf merger using the post-Newtonian framework of \citet{blanchet:1990}.
While a typical merger event has a characteristic profile including a chirp and ringdown,
white dwarf collisions are noticeably distinct in their gravitational wave signatures, and
hence may be interesting targets for observation for future low frequency gravitational
wave detectors such as LISA \citep{eLISA}. The strain increases monotonically for a collision until impact,
instead of oscillating back and forth at a characteristic frequency as for mergers.
The main caveat for these events compared to events
such as neutron star mergers is that due to the event being inherently less energetic,
the typical distance scale on which we could observe a white dwarf merger or collision
is the size of our own galaxy, with extragalactic events being unlikely to be
detected, at least with the first generation of space-based gravitational wave
detectors.

@@

\software{BoxLib\footnote{\url{https://github.com/BoxLib-Codes/BoxLib}},
          CASTRO \citep{castro} (https://github.com/BoxLib-Codes/Castro),
          wdmerger \citep{wdmergerI} (\url{https://github.com/BoxLib-Codes/wdmerger}),
          GCC (\url{https://gcc.gnu.org/}),
          python (\url{https://www.python.org/}),
          matplotlib \citep{matplotlib}, \url{http://matplotlib.org/}),
          yt \citep{yt}, \url{http://yt-project.org/})}

### 2016-07-22 papers/wdmerger_III/paper.tex
Updates to papers

It is important to observe that in this formulation
the radial momentum equation is \textit{not} a conserved quantity: it should and
will in general change with time over the course of the simulation. The above form,
absent of source terms, makes this clear. The angular momentum term in 
\autoref{eq:angular-momentum} can be thought of physically as expressing the fact
that a particle with some initial angular momentum will have its radial momentum
increase with time as it moves away from the origin, and can be thought of geometrically
as expressing the curved nature of the underlying coordinates for the radial momentum
\citep{motl:2002}. Note that while the version in \autoref{eq:radial-momentum-rearranged}
is more convenient for our hydrodynamics methodology, the pressure-gradient terms
in the first version are the ones appropriate for a radial field and so no physical meaning
should be ascribed to the pressure term at the end of \autoref{eq:radial-momentum-rearranged}.

>>>>

It is important to observe that in this formulation the radial momentum equation
is \textit{not} a conserved quantity: it should and will in general change with
time over the course of the simulation if the system is not in perfect rotational
equilibrium. The above form, absent of source terms, makes this clear. The angular momentum term in
\autoref{eq:radial-momentum} can be thought of physically as expressing the fact
that a particle with some initial angular momentum will have its radial momentum
increase with time as it moves away from the origin, and can be thought of geometrically
as expressing the curved nature of the underlying coordinates for the radial momentum
\citep{motl:2002}. It is only the presence of a source term like gravity that can keep
such a particle on an orbit at its original radius, and for such a case it may be helpful
to think of the angular momentum source term as an outward centrifugal force that
balances against the inward centripetal force provided by gravity.
